A Hierarchical Bayesian Framework for Inferring the Stellar Obliquity Distribution

### 2023-04-06 src/tex/ms.tex
new figures

Stellar obliquity $\psi$, the angle between a planet's orbital axis and its host star's spin axis, traces the formation and evolution of a planetary system. In transiting exoplanet observations, only the sky-projected stellar obliquity $\lambda$ can be measured. To determine the true stellar obliquity $\psi$, information about the stellar inclination $i_\star$ is also needed.
In this paper, we demonstrate that while a constraint on the stellar inclination is crucial for measuring the obliquity of an individual system, it is not required for robust determination of the population-level stellar obliquity distribution. 
In practice, the constraints on the stellar obliquity distribution are mainly driven by the sky-projected stellar obliquities.
We introduce a flexible, hierarchical Bayesian framework that can be used to infer the stellar obliquity distribution solely from sky-projected stellar obliquities, including stellar inclination measurements when available.
When applying the framework to all exoplanetary systems with measured sky-projected stellar obliquity, we find that the inferred population-level obliquity distribution is unimodal and peaked at zero degrees. Misaligned systems have nearly isotropic stellar obliquities \edits{with no strong clustering near 90$\degr$. Particularly, the misaligned Hot Jupiter population presents a broad stellar obliquity distribution, which may suggest planet-planet scattering after a convergent disk migration or planetary Kozai might be the dominant mechanisms sculpting the population.}

@@

\section{Relations between $\{\lambda, \lowercase{i}_\star\}$ and $\{\psi, \theta\}$}\label{sec:coords}

In this section, we find the expression of $\lambda$ or $i_\star$ in terms of stellar obliquity $\psi$, the azimuthal angle of the stellar spin axis relative to the orbital axis $\theta$, and the orbital inclination $i_{\rm orb}$. In Figure~\ref{fig:coord}, we introduce two coordinate systems that describe the stellar spin axis $\vb{n_\star}$ and its planet's orbital axis $\vb{n_{\rm orb}}$.

In the $\{\psi, \theta\}$ coordinate system, we assume the planetary orbital axis is aligned with the $\vb{z}$-axis. The stellar spin axis can be written as 
\begin{equation}
    \vb{\hat{n}_\star} = \sin{\psi}\cos{\theta}\vu*{x} + \sin{\psi}\sin{\theta}\vu*{y} + \cos{\psi}\vu*{z}
\end{equation}

In the $\{\lambda, i_\star\}$ coordinate system, where the $\vb{x'}$-axis is defined by the observer's line of sight, the stellar spin axis could be written as 
\begin{equation}
    \vb{\hat{n}_\star}' = \cos{i_\star}\vu*{x}' + \sin{i_\star}\sin{\lambda}\vu*{y}' + \sin{i_\star}\cos{\lambda}\vu*{z}'
\end{equation}

To convert the stellar spin vector $\vb{n_\star}$ from $\vb{x,y,z}$ to $\vb{x',y',z'}$, we apply the rotation matrix to the vector by rotating the vector by $\pi/2 - i_{\rm orb}$ around the $\vb{y'}$-axis. 
\begin{equation}
      \vb{n_\star}' = \mqty[\sin{i_{\rm orb}} & 0 & \cos{i_{\rm orb}} \\ 0 & 1 & 0 \\ -\cos{i_{\rm orb}} & 0 & \sin{i_{\rm orb}}] \vb{n_\star}  
\end{equation}

Now we get $\vb{n_\star}'$ in the $\vb{x',y',z'}$ coordinate expressed by $\psi$, $\theta$, and $i_{\rm orb}$.
\begin{equation}
    \begin{split}
    \vb{n_\star}' = (\sin{\psi}\cos{\theta}\sin{i_{\rm orb}}+\cos{\psi}\cos{i_{\rm orb}})\vu*{x}' &\\
    + \sin{\psi}\sin{\theta}\vu*{y}'& \\
    + (-\sin{\psi}\cos{\theta}\cos{i_{\rm orb}}+\cos{\psi}\sin{i_{\rm orb}})\vu*{z}'&
    \end{split}
\end{equation}

Use the equation, we find the relation between $\lambda$ or $i_\star$ and $\psi$, $\theta$, and $i_{\rm orb}$.

First, comparing the $\vb{x'}$ terms, we find
\begin{equation}
    i_\star = \cos[-1](\sin{\psi}\cos{\theta}\sin{i_{\rm orb}}+\cos{\psi}\cos{i_{\rm orb}}).
\end{equation}

Dividing the $\vb{y'}$ term by the $\vb{z'}$ term, we get
\begin{equation}
    \lambda = \tan[-1](\frac{\sin{\psi}\sin{\theta}}{-\sin{\psi}\cos{\theta}\cos{i_{\rm orb}}+\cos{\psi}\sin{i_{\rm orb}}}).
\end{equation}

@@

With the design, the hyperparameter and parameter priors are the following:
\begin{align}
    w_{0,1} &\sim {\rm Dirichlet}(1, 1) \nonumber\\
    a_0 &\sim \mathcal{U}(0, 50) \nonumber\\
    b_0 &\sim \mathcal{U}(0, 1) \nonumber\\
    a_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    b_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    \cos{i}_{\star,n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \cos{i}_{{\rm orb},n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \lambda_{\star,n} &\sim \mathcal{U}(-\pi, \pi).
\end{align}
The likelihood functions follow:
\begin{align}
    \mathcal{L}(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    \mathcal{L}(i_{\rm orb}) &\sim \prod_{i=1}^N\mathcal{N}(\hat{i}_{{\rm orb},n}, \sigma_{\hat{i}_{{\rm orb},n}}) \nonumber\\
    \mathcal{L}(i_\star) &\sim \prod_{i=1}^N\mathcal{N}({\rm Obs}_{\star,n}, \sigma_{{\rm Obs}_{\star,n}}).
\end{align}
If $\gamma_{\star, n}$ is available, we construct Normal distributions with means and standard deviations from measurements.
Here we use uniform hyperpriors for the Beta distribution. However, it is important to note that when applying the framework to a small sample size with $N \lesssim 50$, the choice of hyperpriors could impact the inferred distribution \citep{Nagpal22}. To ensure the robustness of the inferred distributions in such cases, it is crucial to test their sensitivity to different hyperpriors.

@@

>>>>

Stellar obliquity $\psi$, the angle between a planet's orbital axis and its host star's spin axis, traces the formation and evolution of a planetary system. In transiting exoplanet observations, only the sky-projected stellar obliquity $\lambda$ can be measured. To determine the true stellar obliquity $\psi$, information about the stellar inclination $i_\star$ is also needed.
In this paper, we demonstrate that while a constraint on the stellar inclination is crucial for measuring the obliquity of an individual system, it is not required for robust determination of the population-level stellar obliquity distribution. 
In practice, the constraints on the stellar obliquity distribution are mainly driven by the sky-projected stellar obliquities.
We introduce a flexible, hierarchical Bayesian framework that can be used to infer the stellar obliquity distribution solely from sky-projected stellar obliquities, including stellar inclination measurements when available.
When applying the framework to all exoplanetary systems with measured sky-projected stellar obliquity, we find that the inferred population-level obliquity distribution is unimodal and peaked at zero degrees. Misaligned systems have nearly isotropic stellar obliquities \edits{with no strong clustering near 90$\degr$. 
Furthermore, the stellar obliquity distribution of misaligned Hot Jupiters presents a broad distribution from $0\degr$ to $180\degr$, preferring candidate Hot Jupiter formation mechanisms, such as planet-planet scattering after a convergent disk migration or planetary Kozai, that could produce a broad distribution.}

@@

\section{Coordinate Setup and Transformation}\label{sec:coords}

In this section, we find the expression of $\lambda$ or $i_\star$ in terms of stellar obliquity $\psi$, the azimuthal angle of the stellar spin axis relative to the orbital axis $\theta$, and the orbital inclination $i_{\rm orb}$. In Figure~\ref{fig:coord}, we introduce two coordinate systems that describe the stellar spin axis $\vb{n_\star}$ and its planet's orbital axis $\vb{n_{\rm orb}}$. The setup is similar to the coordinate systems used in \cite{Fabrycky09} but has different coordinate orientations and definitions of the azimuthal angle of the stellar spin.

The angle of primary interest is the angle between the planet's orbital axis and its host star's spin axis $\psi$. To describe the angle in 3$D$, we introduce another angle, the azimuthal angle of the stellar spin around the orbital axis $\theta$. As shown in the left panel in Figure~\ref{fig:coord}, in the $\{\psi, \theta\}$ coordinate system, we put the planetary orbital axis on the $\vb{z}$-axis. The stellar spin axis, described by $\psi$ and $\theta$, can be written as 
\begin{equation}
    \vb{\hat{n}_\star} = \sin{\psi}\cos{\theta}\vu*{x} + \sin{\psi}\sin{\theta}\vu*{y} + \cos{\psi}\vu*{z}
\end{equation}

Next, we introduce an observer-oriented coordinate system. Since we neither know the projected orbital axis in the sky plane nor the projected stellar spin axis, we conveniently set the projected orbital axis to the $\vb*{z}$-axis. In the $\{\lambda, i_\star\}$ coordinate system, shown in the right panel of Figure~\ref{fig:coord}, we point the $\vb{x'}$-axis to the observer. The stellar spin axis could be written as 
\begin{equation}\label{eqn:nstar1}
    \vb{\hat{n}_\star}' = \cos{i_\star}\vu*{x}' + \sin{i_\star}\sin{\lambda}\vu*{y}' + \sin{i_\star}\cos{\lambda}\vu*{z}'
\end{equation}

To transform the stellar spin vector $\vb{\hat{n}_\star}$ from the $\vb*{xyz}$ coordinate to the $\vb*{x'y'z'}$ coordinate, we rotate the $\vb*{xyz}$ coordinate by an angle $\pi/2 - i_{\rm orb}$ about the $\vb{y'}$-axis (the same as $\vb{y}$-axis). 
Applying the rotation matrix such that
\begin{equation}
      \vb{\hat{n}_\star}' = \mqty[\sin{i_{\rm orb}} & 0 & \cos{i_{\rm orb}} \\ 0 & 1 & 0 \\ -\cos{i_{\rm orb}} & 0 & \sin{i_{\rm orb}}] \vb{\hat{n}_\star},
\end{equation}
we can now find the expression of $\vb{\hat{n}_\star}'$ in the $\vb*{x'y'z'}$ coordinate in terms of $\psi$, $\theta$, and $i_{\rm orb}$:
\begin{equation}\label{eqn:nstar2}
    \begin{split}
    \vb{\hat{n}_\star}' = (\sin{\psi}\cos{\theta}\sin{i_{\rm orb}}+\cos{\psi}\cos{i_{\rm orb}})\vu*{x}' &\\
    + \sin{\psi}\sin{\theta}\vu*{y}'& \\
    + (-\sin{\psi}\cos{\theta}\cos{i_{\rm orb}}+\cos{\psi}\sin{i_{\rm orb}})\vu*{z}'&.
    \end{split}
\end{equation}

Comparing Equation~(\ref{eqn:nstar2}) with Equation~(\ref{eqn:nstar1}), we find the expression of $\lambda$ or $i_\star$ in terms of $\psi$, $\theta$, and $i_{\rm orb}$.
First, from the $\vb{\hat{x}'}$ terms, we derive
\begin{equation}
    i_\star = \cos[-1](\sin{\psi}\cos{\theta}\sin{i_{\rm orb}}+\cos{\psi}\cos{i_{\rm orb}}).
\end{equation}
Next, dividing the $\vb{\hat{y}'}$ terms by the $\vb{\hat{z}'}$ terms, we find
\begin{equation}
    \lambda = \tan[-1](\frac{\sin{\psi}\sin{\theta}}{-\sin{\psi}\cos{\theta}\cos{i_{\rm orb}}+\cos{\psi}\sin{i_{\rm orb}}}).
\end{equation}

@@

\edits{In this framework, we limit $i_{\rm orb}$ from $0\degr$ to $90\degr$. To preserve the symmetry, we allow $i_{\rm orb}$ to vary from $0\degr$ to $180\degr$.}
With the design, the hyperparameter and parameter priors are the following:
\begin{align}
    w_{0,1} &\sim {\rm Dirichlet}(1, 1) \nonumber\\
    p_{0,1} &\sim \mathcal{U}(0, 1) \nonumber\\
    q_{0,1} &\sim \mathcal{U}(0, 1) \nonumber\\
    \theta_n &\sim \mathcal{U}(0, \pi)  \nonumber\\
    \cos{i}_{{\rm orb},n} &\sim \mathcal{U}(0, 1).
\end{align}
If $\gamma_{\star, n}$ is available, we construct Normal distributions with means and standard deviations from measurements.

The likelihood functions follow:
\begin{align}
    \mathcal{L}(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    \mathcal{L}(i_{\rm orb}) &\sim \prod_{i=1}^N\mathcal{N}(\hat{i}_{{\rm orb},n}, \sigma_{\hat{i}_{{\rm orb},n}}) \nonumber\\
    \mathcal{L}(i_\star) &\sim \prod_{i=1}^N\mathcal{N}({\rm Obs}_{\star,n}, \sigma_{{\rm Obs}_{\star,n}}).
\end{align}
Here we use uniform hyperpriors for the Beta distribution. However, it is important to note that when applying the framework to a small sample size with $N \lesssim 50$, the choice of hyperpriors could impact the inferred distribution \citep{Nagpal22}. To ensure the robustness of the inferred distributions in such cases, it is crucial to test their sensitivity to different hyperpriors.

@@

\edits{EDIT TO HERE TODAY}
